---
pagetitle: "Session 4"
title: | 
  | Data Wrangling for EDSDers
  | \vspace{1.5cm} \LARGE\emph{Merging datasets}
author: |
  | 13-16 Nov, 2023
  | Tim Riffe
  | Universidad del País Vasco \& Ikerbasque (Basque Foundation for Science)
date: "16 Nov, 2023"
output:
  pdf_document: default 
  html_document:
    number_sections: yes
    toc: yes
params:
  output_dir: "../EDSD2023data/docs"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=12cm, ]{assets/opik_upv_ikerbasque_logo_strip.pdf}\\[\bigskipamount]}
- \posttitle{\end{center}}
bibliography: references.bib
---



<a href="https://github.com/timriffe/EDSD2023data" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#70B7FD; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Note
This material was written during the 16-11-2023 class session and has not been edited thereafter.

## Merging

Eurostat SRH
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(countrycode)
library(readxl)

health <- read_csv("Data/hlth_silc_02_linear.csv.gz",
                   show_col_types = FALSE)
```

We inspected age categories in the data and found that there are noncontinuous 5-year age groups, and oddly centered, but continuous, 10-year age groups. We take the later. The following vector will help us `filter()` the data to just the age groups we need.
```{r}
# health$age |> unique()
ages_keep <-
     c("Y16-24","Y25-34", "Y35-44", "Y45-54",
       "Y55-64","Y65-74","Y75-84","Y_GE85")

# quick check:
parse_number(ages_keep)
```

```{r}
health2 <- 
  health |> 
  filter(age %in% ages_keep,
         isced11 == "TOTAL",
         levels == "VBAD") |> 
  # select and rename
  select(eurostat_code = geo,
         age_fat = age,
         sex,
         year = TIME_PERIOD,
         prev = OBS_VALUE) |> 
  # some final cleaning and recoding,
  # iso3 very important for joining countries
  mutate(age_fat = parse_number(age_fat),
         iso3 = suppressWarnings(countrycode(eurostat_code,
                            origin = "eurostat",
                            destination = "iso3c")),
         prev = prev / 100) |> 
  filter(!is.na(iso3))
```

Now let's expand the health data to single ages, assuming a step function, this little age equivalence table will help us do so with a join. Here are two ways to construct said table, each requiring a silly adjustment after the fact. Neither is aesthetically pleasing. 
```{r}
single_ages <- 16:85
age_table <-
  tibble(age = single_ages,
         # spend some time experimenting with this line to understand it
         age_fat = single_ages - (single_ages + 5) %% 10  ) |> 
  mutate(age_fat = if_else(age_fat == 15, 16, age_fat))

# or manually
age_table <- tibble(age_fat = c(rep(16, 9),
                                rep(seq(25, 75, by = 10),
                                    each = 10),85),
                    age = 16:85) 
```

We'll use this to expand the 10-year age groups to single ages using a join technique (in the end left or right )


```{r}

health_expanded <- 
  health2 |> 
  right_join(age_table, 
             by = join_by(age_fat),
             relationship = "many-to-many") |> 
  select(-eurostat_code,
         -age_fat)
```

WPP single-age data, scripted download failed due to choppy wifi. Main page: <https://population.un.org/wpp/Download/Standard/Population/>. Following chunk set to `eval = FALSE` but it might run for you in interactive mode if you have good internet, hopefully.
```{r, eval = FALSE}
 download.file(
 "https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/2_Population/WPP2022_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx",
      destfile = "Data/WPP2022_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx")
 download.file(
 "https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/2_Population/WPP2022_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx",
      destfile = "Data/WPP2022_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx")
```

* Some people were able to fully download either the male or female datasets, then we swapped files on usb sticks, old school. The following code was mapped schematically before the data was on hand, then quickly written once the data was loaded.

```{r}
Fem <-
  read_excel("Data/WPP2022_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx",
           skip = 16, # skip lines
           guess_max = 1e5) |> # necessary to detect capture col types correctly
  pivot_longer(`0`:`100+`, 
               names_to = "age", 
               values_to = "pop") |> 
  select(iso3 = `ISO3 Alpha-code`,
         year = Year,
         age,
         pop) |> 
  filter(!is.na(iso3)) |> 
  mutate(sex = "F")

# same again for Males
Male <-
  read_excel("Data/WPP2022_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx",
           skip = 16,
           guess_max = 1e5) |> 
  pivot_longer(`0`:`100+`, 
               names_to = "age", 
               values_to = "pop") |> 
  select(iso3 = `ISO3 Alpha-code`,
         year = Year,
         age,
         pop) |> 
  filter(!is.na(iso3)) |> 
  mutate(sex = "M")
```

The Eurostat health data will take care of the filtering when we do the join, i.e. we don't need to explicitly cut down the country, age, or year sets. Note, if you read in these data using just default column class detection you'll be overwhelmed with warnings. By default it guesses column types by checking the first 1000 values, but this data doesn't reveal all classes within that range. Just increase `guess_max` sot that it gets more info to guess classes. HT Maria for the tip!, my strategy would have been to explicitly state column types, but that would be arduous for this data.

Bind together the UN estimates, group ages 85+
```{r}
wpp <- bind_rows(Fem, Male) |> 
  mutate(age = parse_number(age),
         pop = as.numeric(pop) * 1000,
         age = if_else(age > 85, 85, age)) |> 
  group_by(iso3, sex, year, age) |> 
  summarize(pop = sum(pop), .groups = "drop")

```

We join the Eurostat and WPP data together to create `health_burden`, silly name, but whatever. We choose `inner_join()` because WPP data has no total-sex breakdown (we could aggregate to create this if we wanted), whereas the Eurostat health data has no info under age 16 or for ages beyond 85+. 
```{r}
health_burden <- 
  health_expanded |> 
  inner_join(wpp, by = join_by(iso3, year, sex, age)) |> 
  mutate(vbad_count = prev * pop)
```

Check one pattern, just to make sure as expected.
```{r}
health_burden |> 
  filter(iso3 == "FRA",
         year == 2010) |> 
  ggplot(aes(x = age, y = vbad_count, color = sex, group = sex)) +
  geom_line()
```

Narrator: The class was then scared away from the room by the sounds of dripping / gushing water, and the hilarious assurances of INED staff that "everything is ok".


