---
title: "Session 3 notes"
author: "Tim Riffe"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## reshaping

`pivot_longer()` stacks some variables, usually creating one or more "name" columns and usually one value column. It will almost always result in more rows and less columns.

`pivot_wider()` unstacks data, distributing it over columns in some systematic way, for example spreading years over columns.

```{r}
library(gapminder)

gapminder |> 
  pivot_longer(c(lifeExp, pop, gdpPercap), 
               names_to = "variable", 
               values_to = "value") |> 
  pivot_wider(names_from = "variable", values_from = "value")
```

What if we have some sort of hierarchical structure that's distributed over the columns? How do do `pivot_longer()` to lots of variables at once?

This is the brute force approach that will never fail you; The strategy is to pivot to super long, manage columns (if needed), then pivot wider with just one `names_from` column.
```{r}
gapminder |> 
  pivot_wider(names_from = "year", 
              values_from = c("lifeExp", "pop", "gdpPercap")) |> 
  pivot_longer(lifeExp_1952:gdpPercap_2007, 
               #names_to = c("variable", "year"),
               names_to = "variable_year",
               values_to = "value") |>  #,
               #names_sep = "_")
  separate(variable_year, 
           into = c("variable", "year"),
           sep = "_", 
           convert = TRUE) |> 
  pivot_wider(names_from = "variable", values_from = "value")
```

But there is a way to handle the above `pivot_longer()` scenario in fewer steps. Actually `pivot_longer()` can handle all these steps at once. The trick is to specifiy `names_to` as a vector with two parts: `".value"` stands for the various column names, which need to be extracted from the knarly concatenated names you currently have, and then you use `names_sep` to specify the separator. Hopefully you have such as a separator.

```{r}
gapminder |> 
  pivot_wider(names_from = "year", 
              values_from = c("lifeExp", "pop", "gdpPercap")) |> 
  # a one-liner "pivot_longer() to multiple columns"
  pivot_longer(lifeExp_1952:gdpPercap_2007,
               names_to = c(".value","year"),
               names_sep = "_")
```

# HRS data
















