---
title: "Session 2 notes"
author: "Tim Riffe"
date: "2023-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Flat files

```{r}
# options(timeout = 400)
# data_link <- "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/DVS/natality/Nat2022us.zip"
# 
# download.file(url = data_link, destfile = "Data/blabla")

library(R.utils)
gzip("Data/Nat2022us/Nat2022us.txt")
```

In class, we manually downloaded the file, unpacked it, and then g-zipped it using the above line of code. This all took a while because the file is sort of big. We took some care to delete redundant versions of this file because one day our hard-drive might fill up and we'd then need to hunt it down.

In the handout you'll find a fully scripted way to download uncomplicated files like this (meaning to log-in or other obstacles, such as hiding the hard link to the data).

```{r}
# check what the data looks like:
# readLines("Data/Nat2022us/Nat2022us.txt.gz",2)

# browse the documentation
# month 13 - 14
# sex   475
# mage  75-76
# fage  147-148
# apgar 444-445
# bwt   504-507

library(tidyverse)
?read_fwf

# We're going to use read_fwf() using position info created by fwf_positions()

pos <- fwf_positions(start = c(13,475,75,147,444,504),
                     end = c(14,475,76,148,445,507),
                     col_names = c("month","sex","mage","fage","apgar","bwt"))

B <- read_fwf("Data/Nat2022us/Nat2022us.txt.gz",
         col_positions = pos,
         col_types = "iciiii")
B
```

From the documentation pdf, we selected the columns of interest, noting their positions. We then used the `fwf_positions()` function to make a metadata tibble for us to feed to to `read_fwf()`. The only thing you need to remember in case you need to do something like this in the future, you only need to remember that this kind of data file is called a flat file or fixed width file.

## first data processing thing

Here's how to tabulate from microdata, it's just a matter of counting rows within groups. There's a shorthand trick (`count()`), or the verbose way (`group_by() |> summarize()`). You can just list more items alongside `mage` for cross-tabulations.
```{r}
B |> 
  count(mage)

B |> 
  group_by(mage) |> 
  summarize(n = n())
```

# Example, redisitribute missing father age


The task: redistribute births whose age of father is unknown or unstated (99) according to the distribution of known/stated age of fathers.

In practice, you need to manage to convert the counts of "knowns" into a distribution (meaning that it sums to 1) and multiply that distribution into the total count of unknowns.
```{r}
B |> 
  count(fage) |> 
  mutate(unk = n[fage == 99]) |> 
  filter(fage != 99) |> 
  mutate(prop = n / sum(n)) |>  
  # pull(prop) |> sum() # quick check that sums to 1
  
  # quick plot of the distribution we use:
  # ggplot(aes(x = fage, y = prop)) +
  # geom_line()
  mutate(unk_x = prop * unk,
         births = n + unk_x) # the final adjustment
```







