---
title: "Session 2 notes"
author: "Tim Riffe"
date: "2023-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Flat files

```{r}
# options(timeout = 400)
# data_link <- "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/DVS/natality/Nat2022us.zip"
# 
# download.file(url = data_link, destfile = "Data/blabla")

library(R.utils)
gzip("Data/Nat2022us/Nat2022us.txt")
```

In class, we manually downloaded the file, unpacked it, and then g-zipped it using the above line of code. This all took a while because the file is sort of big. We took some care to delete redundant versions of this file because one day our hard-drive might fill up and we'd then need to hunt it down.

In the handout you'll find a fully scripted way to download uncomplicated files like this (meaning to log-in or other obstacles, such as hiding the hard link to the data).

```{r}
# check what the data looks like:
# readLines("Data/Nat2022us/Nat2022us.txt.gz",2)

# browse the documentation
# month 13 - 14
# sex   475
# mage  75-76
# fage  147-148
# apgar 444-445
# bwt   504-507

library(tidyverse)
?read_fwf

# We're going to use read_fwf() using position info created by fwf_positions()

pos <- fwf_positions(start = c(13,475,75,147,444,504),
                     end = c(14,475,76,148,445,507),
                     col_names = c("month","sex","mage","fage","apgar","bwt"))

B <- read_fwf("Data/Nat2022us/Nat2022us.txt.gz",
         col_positions = pos,
         col_types = "iciiii")
B
```

From the documentation pdf, we selected the columns of interest, noting their positions. We then used the `fwf_positions()` function to make a metadata tibble for us to feed to to `read_fwf()`. The only thing you need to remember in case you need to do something like this in the future, you only need to remember that this kind of data file is called a flat file or fixed width file.

## first data processing thing

Here's how to tabulate from microdata, it's just a matter of counting rows within groups. There's a shorthand trick (`count()`), or the verbose way (`group_by() |> summarize()`). You can just list more items alongside `mage` for cross-tabulations.
```{r}
B |> 
  count(mage)

B |> 
  group_by(mage) |> 
  summarize(n = n())
```

# Example, redisitribute missing father age


The task: redistribute births whose age of father is unknown or unstated (99) according to the distribution of known/stated age of fathers.

In practice, you need to manage to convert the counts of "knowns" into a distribution (meaning that it sums to 1) and multiply that distribution into the total count of unknowns.
```{r}
B |> 
  count(fage) |> 
  mutate(unk = n[fage == 99]) |> 
  filter(fage != 99) |> 
  mutate(prop = n / sum(n)) |>  
  # pull(prop) |> sum() # quick check that sums to 1
  
  # quick plot of the distribution we use:
  # ggplot(aes(x = fage, y = prop)) +
  # geom_line()
  mutate(unk_x = prop * unk,
         births = n + unk_x) # the final adjustment
```

Exercise: redo this re-distribution of unknown father age conditional on mother age. What needs to change in the code to handle this?

First, the only thing that needs to change in order to re-condition this whole operation is to 1. include `mage` in the `count()` and 2. apply a `group_by(mage)` before we get roll'n.
```{r}

Bages <-
  B |> 
  count(fage, mage) |>
  group_by(mage) |> 
  mutate(unk = n[fage == 99]) |> 
  filter(fage != 99) |> 
  mutate(prop = n / sum(n)) |>  
  mutate(unk_x = prop * unk,
         births = n + unk_x)  |> 
  ungroup() 

```

# impute 0s where they belong:

Here we fill the gaps; when we tabulate microdata there will always be unobserved combinations of variables. Sometimes we want those variable combinations represented with 0s. That requires an extra step:
```{r}

Bages |> pull(fage) |> range()
all_fages <- 10:98 # the complete fage vector with no gaps
Bages |> 
  select(mage, fage, births) |> 
  complete(mage, 
           fage = all_fages, 
           fill = list(births = 0))
```

plot it!

```{r}
library(colorspace)
hcl_palettes(plot = TRUE)
Bages |> 
  ggplot(aes(x = fage, y = mage, fill = log(births))) +
  geom_tile() +
  coord_equal() +
  scale_fill_continuous_sequential("Red-Purple") +
  theme_minimal()+
  geom_abline(slope = 1, incercept = 0)
```

# grouping ages

To group ages (of counts), redefine the age bins. Here I use a modulo trick (subtract the remainder after dividing out the desired age interval evenly). Then it's a standard `group_by |> summarize()` operation.
```{r}
# 0:20 - 0:20 %% 5

Bages |> 
  mutate(mage = mage - mage %% 2,
         fage = fage - fage %% 2) |> 
  group_by(mage, fage) |> 
  summarize(births = sum(births))
```

# Exercise

Your task is to create a ridge plot (`ggridges` package) of birth weight, where ridgelines are defined by 5-year age groups of mothers. For each ridgeline there are 2 distributions: one for boy births and the other for girl births. Each distribution needs to sum to 1 because we're interested to see whether the distribution appears to change over age and if there are apparent differences by sex assigned at birth (including imputed).

Data prep that will need to:
- bin mother ages
- tabulate by mage, sex, bwt
- redistribute missing birth weights (9999)
- rescale mage-sex-specific birth weight distributions to sum to 1

```{r}
b_to_plot <-
B |> 
  select(mage, sex, bwt) |> 
  
  # 1. define 5-year age bins
  mutate(mage = mage - mage %% 5) |> 
  
  # 2. perform tabulation
  count(mage, sex, bwt) |> 
  
  # 3. impute 0s where needed
  complete(mage, sex, bwt, fill = list(n = 0)) |> 
  
  # 4. declare groups
  group_by(mage, sex) |> 
  
  # 5. move unk values to new column
  mutate(unk = n[bwt == 9999]) |> 
  
  # 6. remove the unk row
  filter(bwt != 9999) |> 
  
  # 7. perform (conditional) redistribution
  #    and calculate proportions
  mutate(births = n + unk * n / sum(n),
         prop = births / sum(births)) |> 
  ungroup() |> 
  # quick check of distribution: is it spiked on even numbers??
  # group_by(bwt) |> 
  # summarize(births = sum(births)) |> 
  # ggplot(aes(x = bwt,y=births)) +
  # geom_line()
  # 
  
  # 8. I'll group birth weight
  mutate(bwt = bwt - bwt %% 50) |> 
  group_by(mage, sex, bwt) |> 
  summarize(prop = sum(prop), .groups = "drop")

b_to_plot
```

Now plot it:

```{r}
library(ggridges)
b_to_plot |> 
  ggplot(aes(x = bwt, 
             y = factor(mage),
             height = prop,
             fill = sex)) +
  geom_ridgeline(scale = 20,
         alpha = .4) +
  theme_minimal() +
  labs(y = "mothers' age",
       x = "birth weight (grams)",
       title = "Boy births are born heavier than girl births")
```


















